---
title: "RchivalTag Tutorial | Part 5 - Visualizing data in local time/other time zones"
description: |
  In part 2 of the RchivalTag tutorial series, we learned already different ways to visualize depth time series data. In this tutorial, we will address how to illustrate the data in local time (other time zones).
author:
  - name: Robert K. Bauer
    url: http://www2.hawaii.edu/~rkbauer
    affiliation: Hawai'i Institute of Marine Biology <br> University of Hawai'i at Mānoa
    affiliation_url: https://scholar.google.com/citations?hl=en&user=J-0_tdbR2tgC
    orcid_id: 0000-0003-4224-8023
date: 02-18-2021
output:
  distill::distill_article:
    toc: true
    toc_float: true
    self_contained: false
# (additional metadata e.g. description and date, excluded for brevity)
categories:
  - RchivalTag
  - local-time
  - timezone
  - UTC
  - time-series
  - dygraphs
creative_commons: CC BY
---

# Getting started | Requirements

To run this tutorial we will need RchivalTag version >= 0.1.5 and oceanmap version >= 0.13. You can find the newest versions of both packages on [github](https://github.com/rkbauer/)

```{r}
## install or load package
# install.packages("RchivalTag") # from CRAN
# library(xfun)
# install_github("rkbauer/oceanmap") # newest version from github
# install_github("rkbauer/RchivalTag") # newest version from github
library("RchivalTag")

## Package overview and version:
?RchivalTag 
help(package="RchivalTag") ## list of functions
```

# Recap: time sieres data plotting functions in RchivalTag
-  `ggboxplot_DepthTS_by_hour`
-  `plot_DepthTS`
-  `plot_DepthTempTS`
-  `plot_DepthTempTS_resampled`
-  `plot_DepthTempTS_resampled_PDT`
-  `dy_DepthTS`

# Visualizing time series data in local time

Let's assume the fish was tagged in Hawaii. The standard time zone of all RchivalTag figures and calculations is UTC (Universal Time Coordinated) which is 10 hours ahead of the local time zone of our tag (Hawaii Standard Time, HST). To facilitate the interpretation of the results, we might like to visualize the results in the local time. Fortunately, ggboxplot_DepthTS_by_hour and the other time series plotting functions of RchivalTag include an argument to define and convert the UTC time of daytime periods and tagging data to your preferred time zone for visualization:

```{r}
library(dplyr)

ts_file <- system.file("example_files/104659-Series.csv",package="RchivalTag")
ts_df <- read_TS(ts_file)
ts_df <- ts_df %>% mutate(Lon = 55.5828,
                          Lat = 19.8968)

ts_df <- classify_DayTime(ts_df)
ts_df$Serial <- ts_df$DeployID 

```


```{r, collapse=TRUE}
## illustrate DepthTS and daytime periods in HST
print(head(ts_df))
ggboxplot_DepthTS_by_hour(ts_df,ylim=c(0,100),tz = "HST") 
```

Let's add the actual depth records on top of the boxplot (only meaningful in case of few amounts of data, low resolution DepthTS):

```{r, collapse=TRUE}
ggboxplot_DepthTS_by_hour(ts_df,tz = "HST", jitter = T, opacity = 0.1)
```

```{r, collapse=TRUE}
plot_DepthTS(ts_df,xlim = unique(ts_df$date)[2:3],tz = "HST",plot_DayTimePeriods=T)
```

```{r, collapse=TRUE}
dy_DepthTS(ts_df,xlim = unique(ts_df$date)[2:3],tz = "HST")
```


# How to find your local time keys

The OlsonNames-function includes all available time zones in R, some of them are abbreviated, others not. The table below shows the time difference from UTC to the local time of each time zone in hours:

```{r}
library(DT)
tzs <- OlsonNames(tzdir = NULL) # list of available time zones
SysTime <- Sys.time()
UTCTime <- SysTime %>% lubridate::with_tz(tz = "UTC")
df <- data.frame(timezone=tzs, SysTime=as.character(SysTime), UTCTime=UTCTime, stringsAsFactors = F)

tzTime <- apply(df,1, function(x) as.character(lubridate::with_tz(SysTime,tzone=x[1])))
df$tzTime <- lubridate::ymd_hms(tzTime, tz = "UTC")
df$time_diff <- round(as.numeric(df$tzTime-df$UTCTime)/(60*60),2)

df %>% mutate(UTCTime=as.character(UTCTime),
              tzTime=as.character(tzTime)) %>%
  # select(timezone,time_diff) %>%
  rename("UTC vs tzTime (h)"=time_diff) %>% 
    datatable(
      extensions = "Buttons",
      options = list(
        dom = 'Bfrtip',
        buttons = c('excel', "csv")))

```

Let's say we look for the time zone of Tonga, a beautiful Polynesian kingdom in the Pacific:

```{r}
tzs[grep("onga",tzs)]
dy_DepthTS(ts_df,xlim = unique(ts_df$date)[2:3],tz = "Pacific/Tongatapu")

```

