---
title: "oceanmap Tutorial | Part 1 - Plotting landmasks"
description: |
  In this first oceanmap tutorial we will learn different ways to plot or add landmasks to existing figures. 
author:
  - name: Robert K. Bauer
    url: http://www2.hawaii.edu/~rkbauer
    affiliation: Hawai'i Institute of Marine Biology <br> University of Hawai'i at Mānoa
    affiliation_url: https://scholar.google.com/citations?hl=en&user=J-0_tdbR2tgC
    orcid_id: 0000-0003-4224-8023
date: 08-16-2020
bibliography: library.bib
output:
  distill::distill_article:
    toc: true
    toc_float: true
    self_contained: false
# (additional metadata e.g. description and date, excluded for brevity)
categories:
  - oceanmap
  - ggplot
  - plotly
  - raster
---

# Getting started | Requirements
To run this tutorial we will need oceanmap version >= 0.13. You can find the newest version on [github](https://github.com/rkbauer/R_Package_oceanmap)

```{r}
## install or load package
# install.packages("oceanmap") # from CRAN
# library(xfun)
# install_github("rkbauer/oceanmap") # newest version from github
library("oceanmap")

## Package overview and version:
?oceanmap 
help(package="oceanmap") ## list of functions
```

# Introduction
Visualizing data is a crucial step in analyzing and exploring data. During the last
two decades the statistical programming language R has become a major tool for data
analyses and visualization across different fields of science. However, creating figures
ready for scientific publication can be a tricky and time consuming task.
The oceanmap package provides some helpful functions to facilitate and optimize the
visualization of geographic and oceanographic data, such as satellite and bathymetric
data sets. Its major functions are

*   plotmap(), ggplotmap() and gglotmaply() to plot land masks
*   set.colorbar(), to define the colorbar (position, ticks, colormap)
*   v(), an optimized tool for creating image or contour plots from different data formats (matrix, array, raster, netcdf, binary)
*   figure(), to facilitate resizing and saving of figures in diverse formats (jpeg, png, eps, pdf and eps)

These functions were written in a way that they do not require a large amount of their numerous arguments to be specified but still return nice plots. Each of these functions
will be subsequently discussed.
A closely related problem represents the extraction of spatial data at defined positions/areas. Some examples on this will be discussed in an extra section.

# The plotmap-function
Often it is required to add a land mask as geographical reference to an image- or scatter-
plot. To add a landmask it is required to provide information about the region extent.
As for the entire oceanmap-package, the idea of the plotmap-function is to reduce the
user effort when creating figures. To allow a high level of user flexibility plotmap can
produce a may by 5 different ways. By providing the required region extent as:

1. geographical coordinates (longitude and latitude),
2. a raster-object
3. an extent-object
4. a region-keyword
5. or add a land mask to an existing plot

Many additional arguments can passed by the user. The idea of a region-keyword
(option 4) is to use a short character string instead of requiring axes limits. This keyword
feature is also implemented in the v-function and permits here the rapid visualization
of subregions of interest (e.g. from a raster object). Region keywords can be set up by
the add.region-function. The latter function is interactive and also defines the default
colorbar placement that is further used by the v-function.

## 5 different ways to produce a map with plotmap() - examples


## Example 1: plot landmask of the Mediterranean Sea
### a) by using longitude and latitude coordinates:

```{r plotting Mediterranean landmask}
lon <- c(-6, 37)
lat <- c(30, 46)
plotmap(lon=lon, lat=lat, main="Mediterranean Sea")
plotmap(xlim=lon, ylim=lat, main="Mediterranean Sea")
```


### b) plot landmask of the Mediterranean Sea by using an extent-object
```{r plotting Mediterranean landmask via extent object}
library('raster')
ext <- extent(lon, lat)
plotmap(ext, main="Mediterranean Sea") # extent-object
```

### c) plot landmask of the Mediterranean Sea by using a raster-object:
```{r plotting Mediterranean landmask via raster object}
r <- raster(ext)
plotmap(r, main="Mediterranean Sea") # raster-object
```

### d) plot landmask of the Mediterranean Sea by using a region label:
```{r plotting Mediterranean landmask via region label}
# run regions() to check pre-installed region labels
# (we will introduce later how to define new region definitions)
plotmap('med4', main="Mediterranean Sea") # region label for the Mediterranean
```


### e) add landmask to an existing plot:
```{r adding Mediterranean landmask to exsiting plot}
plot(3.7008, 43.4079, xlim=lon, ylim=lat)
plotmap(add=T)
points(3.7008, 43.4079, pch=19)
```

## Example 2: subplots and some additional arguments of plotmap()
```{r, layout="l-body-outset", layout="l-body-outset",fig.height=8}
par(mfrow=c(2, 1))
plotmap('medw4', main="Western Mediterranean Sea",col.bg="darkblue")
plotmap('medw4', main="Western Mediterranean Sea", bwd=3, border='grey', 
        grid=FALSE)
```


# Saving plots with figure()

oceanmap inlcudes its own function to save figures in different file formats:

```{saving plots, r}
do.save <- FALSE ## open a plotting window
figure("Gulf_of_Lions_extended", do.save=do.save, width=5, height=5,type="pdf")
plotmap("lion",col.bg='darkblue' ,grid=FALSE)
close_fig(do.save)
```


<style> div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;} </style>
<div class = "blue">
**Recommendation:**
Note that you can, unlike with the basic map-package, resize figures generated by plotmap
(and v)-function calls. (In fact, only the font and border width size will change when
resizing figures.)

Use dev.size() to find the most appropriate dimensions for your plot and check out figure()
to switch between the plotting window and diverse figure file formats: jpeg, png, eps or pdf.
Formats are selected according to the type statement (e.g. type=jpeg).

**ATTENTION** To save plots with figure(), the device needs to be closed by a separate
function(dev.off() or close_fig(/TRUE)) </div>


## Example 3: resize figure manually and get new figure dimensions:
```{r resize figure, eval=FALSE, message=FALSE, error=FALSE, warning=FALSE}
width <- dev.size() [1]
height <- dev.size() [2]

do.save <- TRUE ## do NOT open a plotting window, but save figure internally

figure("Gulf_of_Lions_extended", do.save=do.save, 
       width=width, height=height, type="pdf")
plotmap("lion",col.bg='darkblue' ,grid=FALSE)
close_fig(do.save)
```

# Drawing maps from West to East or East to West:

Particularly tricky but not that uncommon is the case of drawing maps that cover the
region between the “Western” or "Eastern" hemispheres (i.e. that run from 180°W to
180°E or 180°E to 180°W). Whether plotmap draws a map from the Western to the Eastern
or Eastern to Western hemisphere depends on the order and magnitude of the defined
longitude values. In general, xlim and Jon-values < 0 or > 180 are always treated as Western
longitudes. Accordingly, maps are drawn from East fo West if xlim[1] > xlim[2] and from
West to East if xlim[2] > xlim[1]. Here an easy example:

## Example 4: between hemispheres
```{r between hemispheres, layout="l-body-outset", layout="l-body-outset",fig.width = 10, fig.height = 10}
par(mfrow=c(2,1))
plotmap(lon=c(80, -120), lat=c(-50, 10), main="map from East to West")
plotmap(lon=c(-120, 80), lat=c(-50, 10), main="map from West to East")
```

# ggplotmap and and ggplotmaply

The newest version of oceanmap includes now a gglplot and plotly-version of the base-function plotmap(). You can use gglplotmap() to create land masks, which you can add to existing ggplots.

```{r ggplotmap,preview=TRUE}
ggobj <- ggplotmap("mhi")
ggobj
```

Now let's convert this figure into an interative visualization using the ggplotmaply-function:
(Note that the scale bars of ggplotmap() can not be inlcuded in the plotly visualizations.)

```{r ggplotmaply}
ggplotmaply(ggobj)
```

Although the remaining tutorial will focus on the base-functions of oceanmap, we will address how to use ggplotmap and ggplotmaply with rasterized data in the second oceanmap tutorial.


# Available and new region keywords & definitions

Researchers often need to analyse spatial data of a fixed study region (e.g. the Western Mediterranean Sea). To facilitate plotting of such data, oceanmap comes with a region-keyword feature. The idea of this is to link all information required to produce land mask or image plots to a short keyword (label). Apart from its keyword, each region-definition includes a long version of the label, the spatial extent of the region (longitudes and latitudes), its grid resolution, as well as the default colorbar position and default figure dimensions. Available region-definitions are stored as a data frame in the region_definitions file of the package.

```{r region definitions, layout="l-body-outset", fig.width = 15, fig.height = 18,results="hide"}
data(region_definitions)
head (region_definitions)
region_definitions$label
# ?region_definitions
# figure (width=15,height=15)
par(mfrow=c(8,4))
for(n in region_definitions$label) plotmap(region = n,main=n)
```
## How to create and delete region-definitions with add.region()

The function add.region() can be used to set up new region-definitions (and -keywords),
to store or restore a backup. ATTENTION: Backups are necessary since new region-
definitions will be lost when updating eceanmap! In general, new regions can be supplied
as a one-row data frame, e.g. by modifying existing region definition entries, or as
a second option, through an interactive process. Here an example on how to modify
existing definitions (the interactive way will be discussed on the next page):

## Example 1: Add region by supplying a one-row data.frame  that holds the entire required information
```{r, eval=FALSE}
data(region_definitions) # load region_definitions
lion <- region_definitions[region_definitions$label == 'lion',] # selecting Gulf of Lions region
```

## Example 2: Delete region
```{r, eval=FALSE}
delete.region("junk") # delete junk region
data(region_definitions) # reload region_definitions
region_definitions[,1:9]
```

## Example 3: Create new region-definitions with add.region() - the interactive way

Calling add.region() without or insufficient arguments starts an interactive session that
prompts the user to complete the new region definition. The single steps of this way
listed below:

```{r new region label, eval=FALSE, message=FALSE, error=FALSE, warning=FALSE}
Please define the keyword of the new region, coded as 'label':
Please define the long name of the new region, coded as 'name':

Please define the northern most latitude
(negative values for the southern hemisphere) of the new region, coded as 'latn':

Please define the southern most latitude
(negative values for the southern hemisphere) of the new region, coded as 'lats':

Please define the western most longitude of the new region
(negative values for the western hemisphere), coded as 'lonw':

Please define the western most longitude of the new region
(negative values for the western hemisphere), coded as 'lone':

Please type 'm' if you want to perform colorbar placement by hand (mouse cursor)
or a letter (b, 1, t, r) referring to a side of the plot (bottom, left, top, right).

If 'm' (manual placement) was selected:
Please select the lower left colorbar-position by the mouse cursor.

Please select the upper right colorbar-position by the mouse cursor.

You can resize the window to appropriate size. Try to avoid white space.
Press <Enter> when done.

Please enter the default grid resolution

Press <Enter> to save the new region definition
or any other key to abort the operation
```
